Ecosystem size and complexity dictate riverine biodiversity

### 2021-02-18 manuscript.Rmd
edits: automated edits by visual editor

---
title: "Ecosystem size and complexity dictate riverine biodiversity"
author:
- Akira Terui^[Department of Biology, University of North Carolina at Greensboro, Greensboro, NC 27412, USA]
- Seoghyun Kim\footnotemark[1]
- Christine L. Dolph^[Department of Ecology, Evolution and Behavior, University of
  Minnesota, St. Paul, MN 55108, USA]
- Taku Kadoya^[National Institute for Environmental Studies, Tsukuba, Ibaraki 305-8506, Japan]
- Yusuke Miyazaki^[Shiraume Gakuen College, Kodaira, Tokyo 187-8570, Japan]
output:
  word_document: default
  html_notebook: default
abstract: The species-area relationship, a tendency in which species richness increases with increasing area, is a ubiquitous pattern in ecology[@lomolinoEcologyMostGeneral2000] and has formed the basis of area-based conservation strategies[@desmetUsingSpeciesareaRelationship2004; @neigelSpeciesareaRelationshipsMarine2003]. However, many ecosystems possess a complex spatial structure that cannot be represented by area (the scale-invariant complexity)[@rodriguez-iturbeFractalRiverBasins2001; @turnerLandscapeEcologyTheory2015], and its role in scaling biodiversity is largely unknown. Here, we use theory and empirical analysis of extensive fish community data to show that ecosystem size and complexity jointly regulate biodiversity in rivers. Rivers are a prime example of fractal branching networks where the recurrent merging of diverse streams generates a pronounced self-similarity in complex branching patterns[@rodriguez-iturbeFractalRiverBasins2001]. Ecosystem complexity, which we define as the probability of branching per unit river distance[@teruiMetapopulationStabilityBranching2018; @yeakelSynchronisationStabilityRiver2014], should increase watershed-scale species richness by increasing habitat heterogeneity. Meanwhile, ecosystem size should dictate metacommunity size and total habitat diversity, two factors that regulate watershed-scale species richness[@hubbellUnifiedNeutralTheory2001; @kallimanisHowDoesHabitat2008]. Metacommunity simulations in branching networks predicted biodiversity patterns consistent with our hypothesis. Larger and more branched ecosystems supported greater species richness at the watershed scale, and the importance of spatial difference in species composition was apparent when simulated species dispersal is limited. Cross-watershed comparisons across two geographic regions (the Hokkaido island in Japan and the midwestern US) provided empirical support for the theoretical predictions. In both regions, watershed-scale species richness increased with ecosystem size (watershed area) and complexity (branching probability) through enhanced spatial difference in species composition. Recognizing the dual control of biodiversity should broaden viable options for spatial planning of protected areas or restoration, thereby helping conserve biodiversity from societal demands that threaten it.
bibliography: reference.bib
csl: nature.csl
knit: (function(inputFile, encoding) {
        rmarkdown::render(inputFile, encoding = encoding, output_dir = "document_output")
      })
---

@@

## read gis files
filename <- list.files(path = here::here('/empirical/data_gis'), full.names = T)
wsd_subset <- lapply(filename[str_detect(filename, "wsd_subset")], st_read)
point_subset <- lapply(filename[str_detect(filename, "point_subset")], st_read)
shape <- lapply(filename[str_detect(filename, "shape")], st_read)
shape[[1]] <- st_set_crs(shape[[1]], st_crs(wsd_subset[[1]]))

## read theory file
dat_sim <- read_csv(here::here('/theory/result/result_sim2020-11-10.csv'))

@@

# for bayes factor
source(here::here('/empirical/analysis_lm_bf.R'))

# Hokkaido data
dat_hkd_site <- read_csv(here::here('/empirical/data_out/dat_hkd_site.csv'))

# Midwest data
dat_mw_site <- read_csv(here::here('/empirical/data_out/dat_mw_site.csv')) %>% 
  mutate(Year = format(as.Date(Date, format = '%m/%d/%Y'), "%Y"))

@@

Ecologists have long sought to understand the general drivers of biodiversity. One of the most robust empirical generalizations in ecology is the positive relationship between species richness and area, i.e., the species-area relationship (the SAR)[@lomolinoEcologyMostGeneral2000]. In 1921, Arrhenius[@arrheniusSpeciesArea1921] formulated the SAR as a power-law $S = cA^z$, an equation currently known as the Arrhenius species-area relationship ($S$ is the number of species observed in a given geographic area $A$, $c$ the constant, and $z$ the scaling exponent). Since then, the spatial scaling of species richness has been observed in many taxonomic groups[@triantisIslandSpeciesareaRelationship2012]. The SAR is ubiquitous because multiple mechanisms produce an apparently similar pattern. Larger ecosystems typically support more diverse metacommunities due to increased habitat diversity[@kallimanisHowDoesHabitat2008], larger metacommunity size[@hubbellUnifiedNeutralTheory2001], and/or enhanced colonization dynamics[@lososTheoryIslandBiogeography2009]. Importantly, the SAR provides the foundation for global conservation efforts[@myersBiodiversityHotspotsConservation2000; @desmetUsingSpeciesareaRelationship2004;@neigelSpeciesareaRelationshipsMarine2003]. For example, conservation ecologists have used SAR estimates to design marine and terrestrial protected areas[@desmetUsingSpeciesareaRelationship2004; @neigelSpeciesareaRelationshipsMarine2003], which currently encompass more than 30 million km^2^ globally[@deguignet2014UnitedNations2014].

Many ecosystems, however, possess a complex spatial structure that cannot be represented by area - a dimension referred to as scale-invariant complexity[@turnerLandscapeEcologyTheory2015;@rodriguez-iturbeFractalRiverBasins2001]. Such complexity is evident in branching ecosystems, including rivers, trees, and mountain ranges, to name just a few[@rodriguez-iturbeFractalRiverBasins2001]. Geomorphic or biological processes generate a pronounced self-similarity in complex branching patterns such that the part and the whole look alike[@rodriguez-iturbeFractalRiverBasins2001]. Even though the branching structure is independent of spatial scale, it forms a physical template that dictates habitat diversity and dispersal corridors for living organisms[@grantLivingBranchesPopulation2007]. Limited, but accumulating evidence suggests that classical metapopulation and metacommunity theories cannot predict ecological dynamics driven by branching structure[@carraraDendriticConnectivityControls2012; @faganConnectivityFragmentationExtinction2002; @teruiMetapopulationStabilityBranching2018;@brownDendriticNetworkStructure2010], and this recognition has led to recent developments of spatial theories devoted to complex branching ecosystems[@rinaldoRiverNetworksEcological2018;@petersonModellingDendriticEcological2013]. For example, these studies have highlighted key roles of branching structure in driving local biodiversity patterns, such as increased species richness at merging points of branches[@carraraDendriticConnectivityControls2012]. However, most research has explored the consequences of branching complexity for local community structure[@brownDendriticNetworkStructure2010; @carraraDendriticConnectivityControls2012] or has relied solely on theoretical arguments with limited replications of branching architecture[@holtRoleBranchingMaintenance2018]. At present, we lack a comprehensive evaluation of how branching complexity scales biodiversity at the metacommunity level. Filling this knowledge gap may provide common ground for achieving successful conservation in spatially complex ecosystems, where accelerated species loss threatens the delivery of ecosystem services[@dudgeonFreshwaterBiodiversityImportance2006].

@@

$$
\begin{align}
\mu_{z,k} &= \omega(\rho \mu_{z,k+1}^R + \epsilon_{k}^R) + (1-\omega)(\rho \mu_{z,k+1}^L + \epsilon_{k}^L) \\
\omega &= \frac{A_{k+1}^R}{A_{k+1}^R + A_{k+1}^L}
\end{align}
$$
where $R$ and $L$ denote "right" and "left" branches, respectively, and $A_k$ is the number of upstream habitat patches at longitudinal position $k$ (akin to the upstream watershed area). The parameter $\omega$ takes values of $0-1$ and represents the relative influence of the right tributary given the relative size of joining tributaries. With this expression, larger tributaries have a greater influence on the downstream environment, as observed in natural river networks[@bendaNetworkDynamicsHypothesis2004].

### Metacommunity model
We simulated metacommunity dynamics in branching river networks following the metacommunity framework proposed by Thompson et al.[@thompsonProcessbasedMetacommunityFramework2020]. In this framework, abiotic environmental conditions (density-independent), intra- and interspecific competition (density-dependent), and dispersal regulate a species' realized population growth. Below, we describe the details of our metacommunity model.

@@

$$
N_{ix}(t + 1) \sim Poisson(n_{ix}(t) + I_{ix}(t) - E_{ix}(t))
$$
where $n_{ix}(t)$ is the expected number of individuals given the local community dynamics at time $t$, $I_{ix}(t)$ the expected number of immigrants to patch $x$, and $E_{ix}(t)$ the expected number of emigrants from patch $x$. The realized discrete number of individuals is drawn from a Poisson distribution to account for stochasticity in demographic and dispersal processes[@fronhoferClassicalMetapopulationDynamics2017; @thompsonProcessbasedMetacommunityFramework2020]. Local community dynamics are simulated based on the Beverton-Holt equation[@bevertonDynamicsExploitedFish1957]:

@@

where $r_{ix}(t)$ is the reproductive rate of species $i$ given the environmental condition at patch $x$ and time $t$, $r_{0,i}$ the maximum reproductive rate of species $i$, $K_{x}$ the carrying capacity at patch $x$, $\alpha_{ij}$ the competition coefficient between species $i$ and $j$, and $S$ the number of species in a metacommunity. $K_x$ was expressed as a function of the number of upstream habitat patches $A_x$ ($K_x = 100A_x$) to represent increasing habitat size with increasing upstream watershed area[@altermattRiverNetworkProperties2013; @teruiThreeEcologicalFactors2016]. The parameter $\alpha_{ij}$ represents the strength of interspecific competition relative to that of intraspecific competition: interspecific competition is greater than intraspecific competition if $\alpha_{ij}$ > 1 (intraspecific competition coefficient was set constant as $\alpha_{ii} = 1$). $\alpha_{ij}$ was drawn randomly from a uniform distribution as $\alpha_{ij} \sim Unif(0, \alpha_{max})$. The density-independent reproductive rate $r_{ix}(t)$ is affected by abiotic environments (non-consumable) and determined by the following Gaussian function:

$$
r_{ix}(t) = \delta~r_{0,i}~exp[{-\frac{\{\mu_i-z_x(t)\}^2}{2\sigma^2_{niche,i}}}]
$$
where $\mu_i$ is the optimal environmental value for species $i$, $z_x(t)$ the environmental value at patch $x$ and time $t$, and $\sigma_{niche,i}$ the niche width of species $i$. We introduced the cost of a wide niche by multiplying $\delta$[@chaianunpornEvolutionaryResponsesClimate2015]:

@@

The environmental value at patch $x$ and time $t$, $z_x(t)$, is assumed to follow a multivariate normal distribution as $z_{x}(t) \sim MVN(\boldsymbol{\mu_z}, \boldsymbol{\Omega_z})$, where $\boldsymbol{\mu_z}$ is the vector of mean environmental conditions of habitat patches and $\boldsymbol{\Omega_z}$ is the variance-covariance matrix. The diagonal elements of the variance-covariance matrix $\sigma^2_z$ represent the degree of temporal environmental variation. The off-diagonal elements $\Omega_{xy}$ represent spatial autocorrelation in temporal environmental variation as:

@@

$$
\xi_{ix}(t) = \frac{\sum_{y,y \neq x}E_{iy}(t)~exp(-\theta d_{xy})}{\sum_{x} \sum_{y,y \neq x}E_{iy}(t)~exp(-\theta d_{xy})}
$$
The parameter $\theta$ regulates the dispersal distance of a species that follows an exponential distribution ($\theta^{-1}$ corresponds to the expected dispersal distance). The expected number of immigrants is calculated as $I_{ix}(t) = \xi_{ix}(t)\sum_x^{N_p} E_{ix}$.

### Simulation
We used `r n_scenario` sets of parameter combinations (i.e., simulation scenarios) based on the results of extensive sensitivity analysis (**Supplementary Information**). The parameter combinations comprises four landscape and eight ecological scenarios. Landscape scenarios distinguish spatial patterns of habitat heterogeneity, while ecological scenarios change ecological traits of simulated species. For landscape scenarios, we used four combinations of environmental variation at headwaters ($\sigma_h = 0.01, 1$) and the degree of local environmental noise ($\sigma_l = 0.01, 1$). The inequality of $\sigma_h$ and $\sigma_l$ controls the coupling of branching complexity and habitat heterogeneity. For ecological scenarios, we varied three ecological parameters that are relevant for dispersal and interspecific competition: dispersal distance ($\theta =$ `r unique(dat_sim$theta)`), dispersal probability ($p_d =$ `r unique(dat_sim$p_dispersal)`), and the maximum value of competition coefficients ($\alpha_{max} =$ `r unique(dat_sim$max_alpha)`). This setup results in eight ecological scenarios with different levels of dispersal and competition.

@@

We collected data for watershed characteristics (watershed area [km^2^], branching probability, and mean elevation [m]), climatic variables (mean annual air temperature [degree Celsius] and cumulative precipitation [mm]), and land use (the fractions of forest, urban, and agriculture, and dam density per watershed area [km^-2^]). We used Hydrologically Adjusted Elevations of MERIT Hydro[@yamazakiMERITHydroHighResolution2019] to delineate watershed polygons and river lines (defined as $\ge$ 1 km^2^ in the watershed area), which were separated by either the ocean, large lakes/reservoirs ($\ge$ 10 km^2^ in the areal area), or large rivers ($\ge$ 5000 km^2^ in the watershed area). The ocean and large lakes/reservoirs were extracted from Global 1-second Water Body Map ver 1.0[@yamazakiDevelopmentGlobal90m2015]. Watershed area was estimated based on the delineated watershed polygons. We estimated branching probability by fitting exponential distributions to histograms of branch length (unit: kilometer, river segment between successive confluences or a confluence and the outlet/upstream terminal) as $branch~length \sim Exp(\lambda)$. A derived statistical quantity $1 − e^{−\lambda}$ corresponds to the theoretical branching probability $P_b$. We obtained the mean elevations for each watershed polygon using Hydrologically Adjusted Elevations of MERIT Hydro[@yamazakiMERITHydroHighResolution2019]. We estimated the spatial averages of mean annual air temperature and cumulative precipitation for each watershed (~1 km resolution), which were extracted from WorldClim 1.4[@hijmansVeryHighResolution2005]. We calculated the fractions of forest, urban, and agriculture for each watershed using land use data in 2015 from Copernicus Global Land Service (100-m resolution)[@marcelbuchhornCopernicusGlobalLand2020]. We calculated dam density per watershed area (km^-2^) based on the Global Reservoir and Dam Database, Version 1 (GRanDv1)[@lehnerGlobalReservoirDam2011; @lehnerHighresolutionMappingWorld2011]. We used ArcMap 10.7, SAGA 6.3.0, and the following R packages to perform geospatial analysis: 'RSAGA'[@brenningRSAGASAGAGeoprocessing2018], 'sf'[@pebesmaSimpleFeaturesStandardized2018], 'raster'[@hijmansRasterGeographicData2020], 'rgdal'[@rogerbivandRgdalBindingsGeospatial2020], 'stars'[@edzerpebesmaStarsSpatiotemporalArrays2020], and 'exactextractr'[@danielbastonExactextractrFastExtraction2020].

@@

$$
\begin{align}
\mu_w =& \xi_0 + \xi_1~log_{10}~WA_w + \xi_2~log_{10}~P_{b,w} + \xi_3~region_w +\\
&\xi_4~log_{10}~WA_w \cdot region_w + \xi_5~log_{10}~P_{b,w} \cdot region_w + \sum_k \xi_k x_{k,w}
\end{align}
$$
We compared the performance of these competing models using the Bayes factor[@wagenmakersPracticalSolutionPervasive2007], which gives the relative strength of evidence in favor of the global model (M0) over the region-specific model (M1). We used Bayesian Information Criteria (BIC) of the linear models to approximate the Bayes factor[@wagenmakersPracticalSolutionPervasive2007]:

@@

# Acknowledgements
We are grateful to numerous people who have been involved in the collection of fish community data. We thank Michio Fukushima and Hirokazu Urabe for providing information on the Hokkaido Freshwater Fish Database HFish and the fish monitoring data at protected watersheds in Hokkaido. This research was funded by National Science Foundation through a Water Sustainability and Climate Program (WSC) Observatory grant (EAR‐1209402): REACH (Resilience under Accelerated Change) and through a Division of Environmental Biology (DEB 2015634) and by Environmental Protection Agency through a Water Quality Benefits grant (EPA-G2015-STAR-A1).

# Author contributions
AT conceived of the project. AT performed theoretical and statistical analysis with inputs from TK. AT, SK, CLD, and YM compiled and managed the fish community data set. AT wrote the first draft while SK, CLD, TK, and YM contributed significantly to the final version.

# Data and code availability
Third parties provided data for empirical data analysis; references are provided in the main text and Supplementary Information where appropriate. R functions for the generation of branching networks and the metacommunity simulation are provided as the R package 'mcbrnet' (available at https://github.com/aterui/mcbrnet). Codes for simulations, statistical analysis, figures, and tables are available at XXX.

>>>>

---
title: "Ecosystem size and complexity dictate riverine biodiversity"
author:
- Akira Terui^[Department of Biology, University of North Carolina at Greensboro, Greensboro, NC 27412, USA]
- Seoghyun Kim\footnotemark[1]
- Christine L. Dolph^[Department of Ecology, Evolution and Behavior, University of
  Minnesota, St. Paul, MN 55108, USA]
- Taku Kadoya^[National Institute for Environmental Studies, Tsukuba, Ibaraki 305-8506, Japan]
- Yusuke Miyazaki^[Shiraume Gakuen College, Kodaira, Tokyo 187-8570, Japan]
output:
  html_notebook: default
abstract: The species-area relationship, a tendency in which species richness increases with increasing area, is a ubiquitous pattern in ecology[@lomolinoEcologyMostGeneral2000] and has formed the basis of area-based conservation strategies[@desmetUsingSpeciesareaRelationship2004; @neigelSpeciesareaRelationshipsMarine2003]. However, many ecosystems possess a complex spatial structure that cannot be represented by area (the scale-invariant complexity)[@rodriguez-iturbeFractalRiverBasins2001; @turnerLandscapeEcologyTheory2015], and its role in scaling biodiversity is largely unknown. Here, we use theory and empirical analysis of extensive fish community data to show that ecosystem size and complexity jointly regulate biodiversity in rivers. Rivers are a prime example of fractal branching networks where the recurrent merging of diverse streams generates a pronounced self-similarity in complex branching patterns[@rodriguez-iturbeFractalRiverBasins2001]. Ecosystem complexity, which we define as the probability of branching per unit river distance[@teruiMetapopulationStabilityBranching2018; @yeakelSynchronisationStabilityRiver2014], should increase watershed-scale species richness by increasing habitat heterogeneity. Meanwhile, ecosystem size should dictate metacommunity size and total habitat diversity, two factors that regulate watershed-scale species richness[@hubbellUnifiedNeutralTheory2001; @kallimanisHowDoesHabitat2008]. Metacommunity simulations in branching networks predicted biodiversity patterns consistent with our hypothesis. Larger and more branched ecosystems supported greater species richness at the watershed scale, and the importance of spatial difference in species composition was apparent when simulated species dispersal is limited. Cross-watershed comparisons across two geographic regions (the Hokkaido island in Japan and the midwestern US) provided empirical support for the theoretical predictions. In both regions, watershed-scale species richness increased with ecosystem size (watershed area) and complexity (branching probability) through enhanced spatial difference in species composition. Recognizing the dual control of biodiversity should broaden viable options for spatial planning of protected areas or restoration, thereby helping conserve biodiversity from societal demands that threaten it.
bibliography: reference.bib
csl: nature.csl
knit: (function(inputFile, encoding) {
        rmarkdown::render(inputFile, encoding = encoding, output_dir = "document_output")
      })
---

@@

## read gis files
filename <- list.files(path = here::here('empirical/data_gis'), full.names = T)
wsd_subset <- lapply(filename[str_detect(filename, "wsd_subset")], st_read)
point_subset <- lapply(filename[str_detect(filename, "point_subset")], st_read)
shape <- lapply(filename[str_detect(filename, "shape")], st_read)
shape[[1]] <- st_set_crs(shape[[1]], st_crs(wsd_subset[[1]]))

## read theory file
dat_sim <- read_csv(here::here('theory/result/result_sim2020-11-10.csv'))

@@

# for bayes factor
source(here::here('empirical/analysis_lm_bf.R'))

# Hokkaido data
dat_hkd_site <- read_csv(here::here('empirical/data_out/data_hkd_site.csv'))

# Midwest data
dat_mw_site <- read_csv(here::here('empirical/data_out/data_mw_site.csv')) %>% 
  mutate(Year = format(as.Date(Date, format = '%m/%d/%Y'), "%Y"))

@@

Ecologists have long sought to understand the general drivers of biodiversity. One of the most robust empirical generalizations in ecology is the positive relationship between species richness and area, i.e., the species-area relationship (the SAR)[@lomolinoEcologyMostGeneral2000]. In 1921, Arrhenius[@arrheniusSpeciesArea1921] formulated the SAR as a power-law $S = cA^z$, an equation currently known as the Arrhenius species-area relationship ($S$ is the number of species observed in a given geographic area $A$, $c$ the constant, and $z$ the scaling exponent). Since then, the spatial scaling of species richness has been observed in many taxonomic groups[@triantisIslandSpeciesareaRelationship2012]. The SAR is ubiquitous because multiple mechanisms produce an apparently similar pattern. Larger ecosystems typically support more diverse metacommunities due to increased habitat diversity[@kallimanisHowDoesHabitat2008], larger metacommunity size[@hubbellUnifiedNeutralTheory2001], and/or enhanced colonization dynamics[@lososTheoryIslandBiogeography2009]. Importantly, the SAR provides the foundation for global conservation efforts[@myersBiodiversityHotspotsConservation2000; @desmetUsingSpeciesareaRelationship2004; @neigelSpeciesareaRelationshipsMarine2003]. For example, conservation ecologists have used SAR estimates to design marine and terrestrial protected areas[@desmetUsingSpeciesareaRelationship2004; @neigelSpeciesareaRelationshipsMarine2003], which currently encompass more than 30 million km^2^ globally[@deguignet2014UnitedNations2014].

Many ecosystems, however, possess a complex spatial structure that cannot be represented by area - a dimension referred to as scale-invariant complexity[@turnerLandscapeEcologyTheory2015; @rodriguez-iturbeFractalRiverBasins2001]. Such complexity is evident in branching ecosystems, including rivers, trees, and mountain ranges, to name just a few[@rodriguez-iturbeFractalRiverBasins2001]. Geomorphic or biological processes generate a pronounced self-similarity in complex branching patterns such that the part and the whole look alike[@rodriguez-iturbeFractalRiverBasins2001]. Even though the branching structure is independent of spatial scale, it forms a physical template that dictates habitat diversity and dispersal corridors for living organisms[@grantLivingBranchesPopulation2007]. Limited, but accumulating evidence suggests that classical metapopulation and metacommunity theories cannot predict ecological dynamics driven by branching structure[@carraraDendriticConnectivityControls2012; @faganConnectivityFragmentationExtinction2002; @teruiMetapopulationStabilityBranching2018; @brownDendriticNetworkStructure2010], and this recognition has led to recent developments of spatial theories devoted to complex branching ecosystems[@rinaldoRiverNetworksEcological2018; @petersonModellingDendriticEcological2013]. For example, these studies have highlighted key roles of branching structure in driving local biodiversity patterns, such as increased species richness at merging points of branches[@carraraDendriticConnectivityControls2012]. However, most research has explored the consequences of branching complexity for local community structure[@brownDendriticNetworkStructure2010; @carraraDendriticConnectivityControls2012] or has relied solely on theoretical arguments with limited replications of branching architecture[@holtRoleBranchingMaintenance2018]. At present, we lack a comprehensive evaluation of how branching complexity scales biodiversity at the metacommunity level. Filling this knowledge gap may provide common ground for achieving successful conservation in spatially complex ecosystems, where accelerated species loss threatens the delivery of ecosystem services[@dudgeonFreshwaterBiodiversityImportance2006].

@@

$$
\begin{align}
\mu_{z,k} &= \omega(\rho \mu_{z,k+1}^R + \epsilon_{k}^R) + (1-\omega)(\rho \mu_{z,k+1}^L + \epsilon_{k}^L) \\
\omega &= \frac{A_{k+1}^R}{A_{k+1}^R + A_{k+1}^L}
\end{align}
$$

where $R$ and $L$ denote "right" and "left" branches, respectively, and $A_k$ is the number of upstream habitat patches at longitudinal position $k$ (akin to the upstream watershed area). The parameter $\omega$ takes values of $0-1$ and represents the relative influence of the right tributary given the relative size of joining tributaries. With this expression, larger tributaries have a greater influence on the downstream environment, as observed in natural river networks[@bendaNetworkDynamicsHypothesis2004].

### Metacommunity model

We simulated metacommunity dynamics in branching river networks following the metacommunity framework proposed by Thompson et al.[@thompsonProcessbasedMetacommunityFramework2020]. In this framework, abiotic environmental conditions (density-independent), intra- and interspecific competition (density-dependent), and dispersal regulate a species' realized population growth. Below, we describe the details of our metacommunity model.

@@

$$
N_{ix}(t + 1) \sim Poisson(n_{ix}(t) + I_{ix}(t) - E_{ix}(t))
$$

where $n_{ix}(t)$ is the expected number of individuals given the local community dynamics at time $t$, $I_{ix}(t)$ the expected number of immigrants to patch $x$, and $E_{ix}(t)$ the expected number of emigrants from patch $x$. The realized discrete number of individuals is drawn from a Poisson distribution to account for stochasticity in demographic and dispersal processes[@fronhoferClassicalMetapopulationDynamics2017; @thompsonProcessbasedMetacommunityFramework2020]. Local community dynamics are simulated based on the Beverton-Holt equation[@bevertonDynamicsExploitedFish1957]:

@@

where $r_{ix}(t)$ is the reproductive rate of species $i$ given the environmental condition at patch $x$ and time $t$, $r_{0,i}$ the maximum reproductive rate of species $i$, $K_{x}$ the carrying capacity at patch $x$, $\alpha_{ij}$ the competition coefficient between species $i$ and $j$, and $S$ the number of species in a metacommunity. $K_x$ was expressed as a function of the number of upstream habitat patches $A_x$ ($K_x = 100A_x$) to represent increasing habitat size with increasing upstream watershed area[@altermattRiverNetworkProperties2013; @teruiThreeEcologicalFactors2016]. The parameter $\alpha_{ij}$ represents the strength of interspecific competition relative to that of intraspecific competition: interspecific competition is greater than intraspecific competition if $\alpha_{ij}$ \> 1 (intraspecific competition coefficient was set constant as $\alpha_{ii} = 1$). $\alpha_{ij}$ was drawn randomly from a uniform distribution as $\alpha_{ij} \sim Unif(0, \alpha_{max})$. The density-independent reproductive rate $r_{ix}(t)$ is affected by abiotic environments (non-consumable) and determined by the following Gaussian function:

$$
r_{ix}(t) = \delta~r_{0,i}~exp[{-\frac{\{\mu_i-z_x(t)\}^2}{2\sigma^2_{niche,i}}}]
$$

where $\mu_i$ is the optimal environmental value for species $i$, $z_x(t)$ the environmental value at patch $x$ and time $t$, and $\sigma_{niche,i}$ the niche width of species $i$. We introduced the cost of a wide niche by multiplying $\delta$[@chaianunpornEvolutionaryResponsesClimate2015]:

@@

The environmental value at patch $x$ and time $t$, $z_x(t)$, is assumed to follow a multivariate normal distribution as $z_{x}(t) \sim MVN(\boldsymbol{\mu_z}, \boldsymbol{\Omega_z})$, where $\boldsymbol{\mu_z}$ is the vector of mean environmental conditions of habitat patches and $\boldsymbol{\Omega_z}$ is the variance-covariance matrix. The diagonal elements of the variance-covariance matrix $\sigma^2_z$ represent the degree of temporal environmental variation. The off-diagonal elements $\Omega_{xy}$ represent the spatial autocorrelation in temporal environmental variation as:

@@

$$
\xi_{ix}(t) = \frac{\sum_{y,y \neq x}E_{iy}(t)~exp(-\theta d_{xy})}{\sum_{x} \sum_{y,y \neq x}E_{iy}(t)~exp(-\theta d_{xy})}
$$

The parameter $\theta$ regulates the dispersal distance of a species that follows an exponential distribution ($\theta^{-1}$ corresponds to the expected dispersal distance). The expected number of immigrants is calculated as $I_{ix}(t) = \xi_{ix}(t)\sum_x^{N_p} E_{ix}$.

### Simulation

We used `r n_scenario` sets of parameter combinations (i.e., simulation scenarios) based on the results of extensive sensitivity analysis (**Supplementary Information**). The parameter combinations comprises four landscape and eight ecological scenarios. Landscape scenarios distinguish spatial patterns of habitat heterogeneity, while ecological scenarios change ecological traits of simulated species. For landscape scenarios, we used four combinations of environmental variation at headwaters ($\sigma_h = 0.01, 1$) and the degree of local environmental noise ($\sigma_l = 0.01, 1$). The inequality of $\sigma_h$ and $\sigma_l$ controls the coupling of branching complexity and habitat heterogeneity. For ecological scenarios, we varied three ecological parameters that are relevant for dispersal and interspecific competition: dispersal distance ($\theta =$ `r unique(dat_sim$theta)`), dispersal probability ($p_d =$ `r unique(dat_sim$p_dispersal)`), and the maximum value of competition coefficients ($\alpha_{max} =$ `r unique(dat_sim$max_alpha)`). This setup results in eight ecological scenarios with different levels of dispersal and competition.

@@

We collected data for watershed characteristics (watershed area [km^2^], branching probability, and mean elevation [m]), climatic variables (mean annual air temperature [degree Celsius] and cumulative precipitation [mm]), and land use (the fractions of forest, urban, and agriculture, and dam density per watershed area [km^-2^]). We used Hydrologically Adjusted Elevations of MERIT Hydro[@yamazakiMERITHydroHighResolution2019] to delineate watershed polygons and river lines (defined as $\ge$ 1 km^2^ in the watershed area), which were separated by either the ocean, large lakes/reservoirs ($\ge$ 10 km^2^ in the areal area), or large rivers ($\ge$ 5000 km^2^ in the watershed area). The ocean and large lakes/reservoirs were extracted from Global 1-second Water Body Map ver 1.0[@yamazakiDevelopmentGlobal90m2015]. Watershed area was estimated based on the delineated watershed polygons. We estimated branching probability by fitting exponential distributions to histograms of branch length (unit: kilometer, river segment between successive confluences or a confluence and the outlet/upstream terminal) as $branch~length \sim Exp(\lambda)$. A derived statistical quantity $1 − e^{−\lambda}$ corresponds to the theoretical branching probability $P_b$. We obtained the mean elevations for each watershed polygon using Hydrologically Adjusted Elevations of MERIT Hydro[@yamazakiMERITHydroHighResolution2019]. We estimated the spatial averages of mean annual air temperature and cumulative precipitation for each watershed (\~1 km resolution), which were extracted from WorldClim 1.4[@hijmansVeryHighResolution2005]. We calculated the fractions of forest, urban, and agriculture for each watershed using land use data in 2015 from Copernicus Global Land Service (100-m resolution)[@marcelbuchhornCopernicusGlobalLand2020]. We calculated dam density per watershed area (km^-2^) based on the Global Reservoir and Dam Database, Version 1 (GRanDv1)[@lehnerGlobalReservoirDam2011; @lehnerHighresolutionMappingWorld2011]. We used ArcMap 10.7, SAGA 6.3.0, and the following R packages to perform geospatial analysis: 'RSAGA'[@brenningRSAGASAGAGeoprocessing2018], 'sf'[@pebesmaSimpleFeaturesStandardized2018], 'raster'[@hijmansRasterGeographicData2020], 'rgdal'[@rogerbivandRgdalBindingsGeospatial2020], 'stars'[@edzerpebesmaStarsSpatiotemporalArrays2020], and 'exactextractr'[@danielbastonExactextractrFastExtraction2020].

@@

$$
\begin{align}
\mu_w =& \xi_0 + \xi_1~log_{10}~WA_w + \xi_2~log_{10}~P_{b,w} + \xi_3~region_w +\\
&\xi_4~log_{10}~WA_w \cdot region_w + \xi_5~log_{10}~P_{b,w} \cdot region_w + \sum_k \xi_k x_{k,w}
\end{align}
$$

We compared the performance of these competing models using the Bayes factor[@wagenmakersPracticalSolutionPervasive2007], which gives the relative strength of evidence in favor of the global model (M0) over the region-specific model (M1). We used Bayesian Information Criteria (BIC) of the linear models to approximate the Bayes factor[@wagenmakersPracticalSolutionPervasive2007]:

@@

# Acknowledgements

We are grateful to numerous people who have been involved in the collection of fish community data. We thank Michio Fukushima and Hirokazu Urabe for providing information on the Hokkaido Freshwater Fish Database HFish and the fish monitoring data at protected watersheds in Hokkaido. This research was funded by National Science Foundation through a Water Sustainability and Climate Program (WSC) Observatory grant (EAR‐1209402): REACH (Resilience under Accelerated Change) and through a Division of Environmental Biology (DEB 2015634) and by Environmental Protection Agency through a Water Quality Benefits grant (EPA-G2015-STAR-A1).

# Author contributions

AT conceived of the project. AT performed theoretical and statistical analysis with inputs from TK. AT, SK, CLD, and YM compiled and managed the fish community data set. AT wrote the first draft while SK, CLD, TK, and YM contributed significantly to the final version.

# Data and code availability

Third parties provided data for empirical data analysis; references are provided in the main text and Supplementary Information where appropriate. R functions for the generation of branching networks and the metacommunity simulation are provided as the R package 'mcbrnet' (available at <https://github.com/aterui/mcbrnet>). Codes for simulations, statistical analysis, figures, and tables are available at XXX.
